Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2023-12-22 doc/JAIR/jair-paper.tex
Correct bug in structured proof description

This paper is an extended version of one published at the
2023 Conference on the Theory and Application of Boolean Satisfiability~\cite{bryant:sat:2023}.

@@

  We highlight some special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

@@

As notation,
let $\psi$ be a subset of the clauses in $\inputformula$.
For partial assignment
$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
obtained from $\psi$ by: i) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
ii) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and iii) eliminating any duplicate or tautological clauses.
In doing these simplifications, we also track the \emph{provenance}
of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
those clauses $C' \in \psi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\prov$ to any simplified formula
$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a dec-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$ with a single RUP step,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Observe that
all of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

The combination of these two optimization guarantees that i) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results show that these
optimizations yield substantial benefits.

>>>>

This paper is an extended version of one published at the 2023
Conference on the Theory and Application of Boolean
Satisfiability~\cite{bryant:sat:2023}.  It provides greater detail of
the algorithms, the formal verification, and the experimental results.

@@

  We highlight some special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to
  variables.  A formula is in \emph{conjunctive normal form} (CNF)
  when (i) it is in negation normal form, and (ii) sum is applied only
  to literals, and (iii) there is a single product operation over all
  of the sums.  A CNF formula can be represented as a set of
  \emph{clauses}, each of which is a set of literals.  Each clause
  represents the sum of the literals, and the formula is the product
  of its clauses.  We use set notation to reference the clauses within
  a formula and the literals within a clause.  A clause consisting of
  a single literal is referred to as a \emph{unit} clause and the
  literal as a \emph{unit} literal.  This literal must be assigned
  value $1$ by any satisfying assignment of the formula.

@@

As notation,
let $\psi$ be a subset of the clauses in $\inputformula$.
For partial assignment
$\passign$, the expression  $\simplify{\psi}{\passign}$ denotes the set of clauses $\gamma$
obtained from $\psi$ by: (i) eliminating any
clause containing a literal $\lit$ such that $\passign(\lit) = 1$,
(ii) for the remaining clauses eliminating those literals $\lit$ for
which $\passign(\lit) = 0$, and (iii) eliminating any duplicate or tautological clauses.
In doing these simplifications, we also track the \emph{provenance}
of each simplified clause $C$, i.e., which of the (possibly multiple) input clauses simplified to become $C$.
More formally, for $C \in \simplify{\psi}{\passign}$, we let $\prov_{\passign}(C, \psi)$ denote
those clauses $C' \in \psi$, such that
$C' \subseteq C \cup \bigcup_{\lit \in \passign} \obar{\lit}$.
We then extend the definition of $\prov$ to any simplified formula
$\gamma$ as $\prov_{\passign}(\gamma, \psi) = \bigcup_{C \in \gamma} \prov_{\passign}(C, \psi)$.

@@

The process for generating such a proof depends on the form of node $\nodeu$:
\begin{enumerate}
\item If $\nodeu$ is a literal $\lit'$, then the formula
  $\simplify{\psi}{\passign}$ must consist of the single unit clause
  $C = \{\lit'\}$, such that any $C' \in \prov_{\passign}(C, \psi)$ must have $C' \subseteq \{ \lit' \} \cup\, \bigcup_{\lit \in \passign} \obar{\lit}$.
  Any of these can
  serve as the target clause.
\item If $\nodeu$ is a sum node with children $\nodeu_1$ and $\nodeu_0$,
  then, since the node originated from a dec-DNNF graph, there must be
  some variable $x$ such that either $\nodeu_1$ is a literal node for $x$ or $\nodeu_1$ is a
  product node containing a literal node for $x$ as a child.  In either case, we
  recursively call $\validate(\nodeu_1, \passign \cup \{ x \}, \psi)$.
  This will cause the addition of the target clause
  $u_1 \lor \obar{x} \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
Similarly, either $\nodeu_0$ is a literal node for $\obar{x}$ or $\nodeu_0$ is a product node containing a literal node for $\obar{x}$ as
  a child.  In either case, we recursively call $\validate(\nodeu_0, \passign \cup \{ \obar{x} \}, \psi)$,
  causing the addition of the target clause
  $u_0 \lor x \lor \bigvee_{\lit \in \passign} \obar{\lit}$.
  These recursive results can be combined with the second and third defining clauses for $\nodeu$
(see Table~\ref{tab:defining}(B))
  to generate the target clause for $\nodeu$, requiring at most two RUP steps.
\item If $\nodeu$ is a product node, then we can divide its children
  into a set of literal nodes $\lambda$ and a set of nonterminal nodes $\nodeu_1, \nodeu_2, \ldots, \nodeu_k$.
  \begin{enumerate}
    \item For each literal
  $\lit \in \lambda$, we must prove that any total assignment $\alpha$, such that
  $\passign \subset \alpha$ has $\alpha(\lit) = 1$.  In some
  cases, this can be done by simple Boolean constraint propagation (BCP).
  In other cases, we must prove that the formula
  $\simplify{\psi}{\passign \cup \{\obar{\lit}\}}$ is unsatisfiable.  We
  do so by writing the formula to a file, invoking a proof-generating
  SAT solver, and then converting the generated unsatisfiability proof
  into a sequence of clause additions in the CPOG file.
  (The solver is constrained to only use RUP inference rules, preventing it from introducing extension variables.)
\item For a single nonterminal child ($k = 1$), we recursively call
  $\validate \left(\nodeu_1, \passign \cup \lambda, \psi\right)$.
\item For multiple nonterminal children ($k > 1$),
  it must be the case that the clauses in
  $\gamma = \simplify{\psi}{\passign \cup \lambda}$ can be partitioned into $k$ subsets
  $\gamma_1, \gamma_2, \ldots, \gamma_k$ such that $\dependencyset(\gamma_i)
  \cap \dependencyset(\gamma_j) = \emptyset$ for $1 \leq i < j \leq k$,
  and we can match each node $\nodeu_i$ to subset $\gamma_i$ based on its
  literals.
  For each $i$ such that $1 \leq i \leq k$, let $\psi_i = \prov_{\passign}(\gamma_i, \psi)$, i.e., those input clauses in $\psi$ that, when simplified, became clause partition $\gamma_i$.
  We recursively call
  $\validate \left(\nodeu_i, \passign \cup \lambda, \psi_i\right)$.
\end{enumerate}
  We then generate the target clause for node $\nodeu$ with a single RUP step,
creating the hint by combining the results from the BCP and SAT calls for
  the literals, the recursively computed target clauses, and all but
  the first defining clause for node $\nodeu$
(see Table~\ref{tab:defining}(A)).
\end{enumerate}
Observe that
all of these steps involve a polynomial number of
operations per recursive call, with the exception of those that call
a SAT solver to validate a literal.

@@

The combination of these two optimization guarantees that (i) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and (ii) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results show that these
optimizations yield substantial benefits.
